Effectful Semantics for Substructural SSA

### 2023-11-14 pldi24/paper.tex
Tap

In particular, the judgement \(\issubst{\gamma}{\Theta}{\Gamma}\) says that
\(\gamma\), which is taken to be a map from variable names to expressions, is a
valid substitution taking \(\Theta\) to \(\Gamma\). What this means is that, for
each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted
as a \textit{pure} term of type \(A^q\) in a subcontext of \(\Theta\), such that
when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
in \(\Theta\) are used up: this is described formally via the typing rules
\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).

\TODO{some text about substitution operations}

@@

\TODO{shuffle subst/lsubst and then ops?}

>>>>

In particular, the judgement \(\issubst{\gamma}{\Theta}{\Gamma}\) says that
\(\gamma\), which is taken to be a map from variable names to expressions, is a
valid substitution taking \(\Theta\) to \(\Gamma\). What this means is that, for
each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted
as a \textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such
that when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine
variables in \(\Theta\) are used up: this is described formally via the typing
rules \rle{subst-nil}, which says that any function \(\gamma\) takes the empty
context to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).

Given such a substitution \(\gamma\), there are two ways to use it. The first is
standard capture-avoiding substitution, which we denote \([\gamma]e\) for
expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for label-sets, as is
standard. The second is to use the substitution to generate a set of
let-bindings for a context \(\Gamma\), written
\(\exprletsubst{\substctx{\gamma}{\Gamma}}{e}\) for expressions and
\(\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}\) for blocks, as defined in Figure
\ref{fig:ssa-subst-ops}. Note that this operation cannot be easily defined for
label-sets. In Section \ref{sec:semantics}, we will show that this is
semantically equivalent to regular capture-avoiding substitution, due to the
fact that all terms in a substitution are central \textit{and} that only affine
computations can be used in an expression substituting for an affine variable,
and only relevant computationbs can be used in an expression substituting
for a relevant variable.

We can now state our substitution theorem for terms in a straightforward way: if
\(\substctx{\gamma}{\Theta}{\Gamma}\) and \(\hasty{\Gamma}{p}{e}{A}{q}\), then
\(\hasty{\Theta}{p}{[\gamma]e}{A}{q}\) and
\(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{e}}{A}{q}\).
Unfortunately, the substitution theorem for blocks is a little more complicated:
the naive choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may contain
captured variables from \(\Gamma\) that are no longer available if we are typing
\([\gamma]t\) in the context of \(\Theta\). To solve this problem, we need to
introduce the concept of a \textit{label-substitution} \(\mc{L}\) \todo{rest}

@@
